---
title: "OTTERS: Omnibus TWAS using multiple RSS methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OTTERS: Omnibus TWAS using multiple RSS methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

The OTTERS pipeline (Zhang et al. 2024) performs TWAS in two stages:

1. **Stage I** (`otters_weights`): Train eQTL weights using multiple
   summary-statistics-based (RSS) methods — P+T, lassosum, PRS-CS, SDPR,
   and any additional learner following the `*_weights(stat, LD, ...)` convention.
2. **Stage II** (`otters_association`): Compute per-method TWAS z-scores via the
   FUSION formula and combine p-values across methods using ACAT.

This vignette demonstrates the pipeline using simulated data. For real data,
you would load eQTL summary statistics and LD matrices using pecotmr's
data loading infrastructure:

- `load_rss_data()` — load eQTL summary statistics from file
- `load_LD_matrix()` — load LD correlation matrix from PLINK2 genotypes or
  pre-computed blocks
- `compute_LD()` — compute LD from a genotype matrix (with optional shrinkage
  parameter for lassosum-style regularization)
- `ld_loader()` — create a lazy loader for genome-wide block-by-block processing
- `rss_basic_qc()` — allele QC, strand flip handling, and variant matching

## Simulate data

We simulate a gene region with 50 SNPs, 4 of which are causal eQTLs.
The gene's predicted expression has a causal effect on a GWAS trait.

```{r simulate}
library(pecotmr)
set.seed(2024)

# Parameters
n_eqtl <- 500    # eQTL study sample size
n_gwas <- 50000  # GWAS sample size
p <- 50           # SNPs in gene region

# Simulate genotype matrix (reference panel) with realistic LD
X <- matrix(rbinom(n_eqtl * p, 2, 0.3), nrow = n_eqtl)
R <- cor(X)

# True eQTL effects (sparse: 4 causal SNPs out of 50)
beta_eqtl <- rep(0, p)
causal <- c(5, 15, 25, 40)
beta_eqtl[causal] <- c(0.3, -0.25, 0.2, -0.15)

# Simulate gene expression and compute eQTL z-scores
expression <- X %*% beta_eqtl + rnorm(n_eqtl)
eqtl_z <- as.vector(cor(expression, X)) * sqrt(n_eqtl)
eqtl_sumstats <- data.frame(z = eqtl_z)

# Simulate GWAS z-scores (gene has causal effect on trait)
beta_gwas_on_trait <- 0.05  # gene-level effect on trait
gwas_z <- as.numeric(R %*% (beta_eqtl * beta_gwas_on_trait * sqrt(n_gwas))) +
          rnorm(p)

cat("Causal eQTL SNPs:", causal, "\n")
cat("True eQTL effects:", beta_eqtl[causal], "\n")
cat("Gene effect on trait:", beta_gwas_on_trait, "\n")
```

## Stage I: Train eQTL weights

`otters_weights()` trains weights using multiple methods simultaneously.
By default it runs P+T (at p < 0.001 and p < 0.05), lassosum, PRS-CS,
and SDPR. You can customize which methods to include and pass
method-specific arguments.

```{r train, message=FALSE}
weights <- otters_weights(
  sumstats = eqtl_sumstats,
  LD = R,
  n = n_eqtl,
  methods = list(
    lassosum_rss = list(),
    prs_cs = list(n_iter = 1000, n_burnin = 500, thin = 5, seed = 42)
  ),
  p_thresholds = c(0.001, 0.05)
)

# Summary of trained weights
for (nm in names(weights)) {
  w <- weights[[nm]]
  cat(sprintf("%-18s: %d non-zero weights (out of %d)\n",
              nm, sum(w != 0), length(w)))
}
```

### Correlation with true eQTL effects

```{r weight_quality}
for (nm in names(weights)) {
  w <- weights[[nm]]
  if (any(w != 0)) {
    cat(sprintf("%-18s: cor(weights, truth) = %.4f\n",
                nm, cor(w, beta_eqtl)))
  }
}
```

## Stage II: TWAS association testing

`otters_association()` computes per-method TWAS z-scores using the
FUSION formula $Z_{TWAS} = w^T z / \sqrt{w^T R w}$ and combines
p-values via ACAT.

```{r test}
twas_results <- otters_association(
  weights = weights,
  gwas_z = gwas_z,
  LD = R,
  combine_method = "acat"
)

print(twas_results)
```

The `ACAT_combined` row aggregates evidence across all methods. This
omnibus test is the main output of the OTTERS framework — it is more
powerful than any single method because different methods capture
different genetic architectures.

## Extending with additional learners

Any function following the `*_weights(stat, LD, ...)` convention can
be added as a learner. For example, to include `mr_ash_rss`:

```{r extend, eval=FALSE}
weights <- otters_weights(
  sumstats = eqtl_sumstats,
  LD = R,
  n = n_eqtl,
  methods = list(
    lassosum_rss = list(),
    prs_cs = list(n_iter = 1000, n_burnin = 500, thin = 5),
    sdpr = list(iter = 1000, burn = 200, thin = 5, verbose = FALSE),
    mr_ash_rss = list(var_y = 1, sigma2_e = 0.5,
                      s0 = c(0, 0.01, 0.1, 0.5), w0 = rep(0.25, 4))
  )
)
```

No code changes are needed — the method is dispatched dynamically via
`do.call(paste0(method_name, "_weights"), ...)`.

## Genome-wide usage with ld_loader()

For genome-wide TWAS across many gene regions, use `ld_loader()` to
avoid loading all LD blocks into memory at once:

```{r genomewide, eval=FALSE}
# Create lazy LD loader from pecotmr metadata
loader <- ld_loader(ld_meta_path = "path/to/ld_meta.tsv",
                    regions = gene_regions)

# Process each gene
for (g in seq_along(gene_regions)) {
  R_g <- loader(g)  # loads LD on demand

  weights_g <- otters_weights(eqtl_sumstats_g, R_g, n_eqtl)
  twas_g <- otters_association(weights_g, gwas_z_g, R_g)
}
```

## References

- Zhang et al. (2024). OTTERS: a powerful TWAS framework leveraging
  summary-level reference data. *Nature Communications*.
- Gusev et al. (2016). Integrative approaches for large-scale
  transcriptome-wide association studies. *Nature Genetics*.
- Mak et al. (2017). Polygenic scores via penalized regression on
  summary statistics. *Genetic Epidemiology*.
- Ge et al. (2019). Polygenic prediction via Bayesian regression and
  continuous shrinkage priors. *Nature Communications*.
- Zhou et al. (2023). SDPR: a statistical method for leveraging
  summary-level data for polygenic risk prediction.
