---
title: "OTTERS: Omnibus TWAS using multiple RSS methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OTTERS: Omnibus TWAS using multiple RSS methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

The OTTERS pipeline (Zhang et al. 2024) performs TWAS in two stages:

1. **Stage I** (`otters_weights`): Train eQTL weights using multiple
   summary-statistics-based (RSS) methods.
2. **Stage II** (`otters_association`): Compute per-method TWAS z-scores
   and combine p-values across methods using ACAT.

This vignette demonstrates the pipeline using simulated data.

**Data loading (not shown here)**: For real data, load eQTL summary
statistics and LD using pecotmr's infrastructure:

- `load_rss_data()` — load eQTL summary statistics from tabix-indexed files
- `load_LD_matrix()` — load LD from PLINK2 pgen/pvar/psam or pre-computed blocks
- `compute_LD()` — compute LD from a genotype matrix, with optional
  `shrinkage` parameter for lassosum-style regularization
- `ld_loader()` — create a lazy loader for genome-wide block-by-block processing
- `rss_basic_qc()` / `allele_qc()` — allele harmonization, strand flip handling

## Simulate data

We simulate a gene region with 50 SNPs, 4 of which are causal eQTLs.
The gene's predicted expression has a causal effect on a GWAS trait.

```{r simulate}
library(pecotmr)
set.seed(2024)

# Parameters
n_eqtl <- 500    # eQTL study sample size
n_gwas <- 50000  # GWAS sample size
p <- 50           # SNPs in gene region

# Simulate genotype matrix (reference panel) with realistic LD
X <- matrix(rbinom(n_eqtl * p, 2, 0.3), nrow = n_eqtl)
R <- cor(X)

# True eQTL effects (sparse: 4 causal SNPs out of 50)
beta_eqtl <- rep(0, p)
causal <- c(5, 15, 25, 40)
beta_eqtl[causal] <- c(0.3, -0.25, 0.2, -0.15)

# Simulate gene expression and compute eQTL z-scores
expression <- X %*% beta_eqtl + rnorm(n_eqtl)
eqtl_z <- as.vector(cor(expression, X)) * sqrt(n_eqtl)
eqtl_sumstats <- data.frame(z = eqtl_z)

# Simulate GWAS z-scores (gene has causal effect on trait)
beta_gwas_on_trait <- 0.05
gwas_z <- as.numeric(R %*% (beta_eqtl * beta_gwas_on_trait * sqrt(n_gwas))) +
          rnorm(p)

cat("Causal eQTL SNPs:", causal, "\n")
cat("True eQTL effects:", beta_eqtl[causal], "\n")
```

## Stage I: Train eQTL weights

### Default methods and parameters

`otters_weights()` runs three methods by default, with parameters matching
the original OTTERS (Zhang et al. 2024):

| Method | Key defaults | Description |
|--------|-------------|-------------|
| **P+T** | `p_thresholds = c(0.001, 0.05)` | Select SNPs by eQTL p-value; weights = marginal z/sqrt(n) |
| **lassosum** | `s = c(0.2, 0.5, 0.9, 1.0)`, `lambda` = 20 values from 1e-4 to 0.1 | L1-penalized regression on summary stats; grid search over shrinkage and penalty |
| **PRS-CS** | `phi = 1e-4` (fixed), `n_iter = 1000`, `n_burnin = 500`, `thin = 5` | Bayesian continuous shrinkage prior; fixed global shrinkage matches OTTERS default |
| **SDPR** | `iter = 1000`, `burn = 200`, `thin = 1` | Dirichlet process mixture; no thinning matches OTTERS default |

These defaults are set in the `methods` argument:

```r
# This is the default — you don't need to type it:
methods = list(
  lassosum_rss = list(),
  prs_cs = list(phi = 1e-4, n_iter = 1000, n_burnin = 500, thin = 5),
  sdpr = list(iter = 1000, burn = 200, thin = 1, verbose = FALSE)
)
```

### Running with defaults

```{r train, message=FALSE}
weights <- otters_weights(
  sumstats = eqtl_sumstats,
  LD = R,
  n = n_eqtl
)

for (nm in names(weights)) {
  w <- weights[[nm]]
  cr <- if (any(w != 0)) round(cor(w, beta_eqtl), 3) else NA
  cat(sprintf("%-18s: %2d non-zero weights, cor(w, truth) = %s\n",
              nm, sum(w != 0), cr))
}
```

## Stage II: TWAS association testing

`otters_association()` computes per-method TWAS z-scores using the
FUSION formula and combines p-values via ACAT:

$$Z_{TWAS} = \frac{w^T z}{\sqrt{w^T R w}}$$

```{r test}
twas_results <- otters_association(
  weights = weights,
  gwas_z = gwas_z,
  LD = R
)
print(twas_results)
```

The `ACAT_combined` row aggregates evidence across all methods using the
Cauchy combination test (Liu & Xie, 2020). This omnibus test is the main
output of the OTTERS framework — it is more powerful than any single method
because different methods capture different genetic architectures.

## Extending with additional learners

Any function following the `*_weights(stat, LD, ...)` convention can be
added as a learner. No code changes needed — methods are dispatched
dynamically via `do.call()`.

### Example: adding mr.ash.rss

`mr_ash_rss_weights()` (from susieR's mixture of regressions with adaptive
shrinkage) is a natural addition — it uses a flexible mixture-of-normals
prior that can capture both sparse and polygenic architectures:

```{r extend, eval=FALSE}
weights <- otters_weights(
  sumstats = eqtl_sumstats,
  LD = R,
  n = n_eqtl,
  methods = list(
    # Original OTTERS methods (defaults shown explicitly)
    lassosum_rss = list(),
    prs_cs = list(phi = 1e-4, n_iter = 1000, n_burnin = 500, thin = 5),
    sdpr = list(iter = 1000, burn = 200, thin = 1, verbose = FALSE),
    # Additional learner: mr.ash.rss
    mr_ash_rss = list(
      var_y = 1,                            # phenotype variance (1 for standardized)
      sigma2_e = 0.5,                       # residual variance
      s0 = c(0, 0.001, 0.01, 0.1, 0.5),    # prior mixture variances
      w0 = rep(0.2, 5)                      # prior mixture weights (uniform)
    )
  )
)
```

The `mr_ash_rss` method uses a mixture-of-normals prior with 5
components (including a point mass at zero), covering a range from
null to large effects. The `s0` grid and `w0` weights can be tuned,
but the defaults above are a reasonable starting point for most eQTL
applications.

### Example: minimal two-method run

For quick analyses or limited compute, you can run just P+T and lassosum:

```{r minimal, eval=FALSE}
weights <- otters_weights(
  sumstats = eqtl_sumstats,
  LD = R,
  n = n_eqtl,
  methods = list(lassosum_rss = list()),
  p_thresholds = c(0.05)
)
```

## Genome-wide usage with ld_loader()

For genome-wide TWAS across many gene regions, use `ld_loader()` to
avoid loading all LD blocks into memory at once:

```{r genomewide, eval=FALSE}
# Create lazy LD loader from pecotmr metadata
loader <- ld_loader(ld_meta_path = "path/to/ld_meta.tsv",
                    regions = gene_regions)

# Process each gene
results_list <- lapply(seq_along(gene_regions), function(g) {
  R_g <- loader(g)
  weights_g <- otters_weights(eqtl_sumstats_list[[g]], R_g, n_eqtl)
  otters_association(weights_g, gwas_z_list[[g]], R_g)
})
all_results <- do.call(rbind, results_list)
```

## Note on variant alignment

`otters_association()` assumes that weights, GWAS z-scores, and the LD
matrix are aligned to the same variant set and allele orientation. For
real data, use `allele_qc()` or `rss_basic_qc()` to harmonize before
calling the pipeline.

## References

- Zhang et al. (2024). OTTERS: a powerful TWAS framework leveraging
  summary-level reference data. *Nature Communications*.
- Gusev et al. (2016). Integrative approaches for large-scale
  transcriptome-wide association studies. *Nature Genetics*.
- Mak et al. (2017). Polygenic scores via penalized regression on
  summary statistics. *Genetic Epidemiology*.
- Ge et al. (2019). Polygenic prediction via Bayesian regression and
  continuous shrinkage priors. *Nature Communications*.
- Zhou et al. (2023). SDPR: a statistical method for leveraging
  summary-level data for polygenic risk prediction.
- Liu & Xie (2020). Cauchy combination test: a powerful test with
  analytic p-value calculation. *JASA*.
